The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth

### 2022-01-14 paper/paper.tex
revised introduction

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{lipsum}
\usepackage{graphicx}

@@

\author{
    Dewi Amaliah
   \\
    Department of Econometrics and Business Statistics \\
    Monash University \\
  Clayton, VIC 3800 \\
  \texttt{\href{mailto:dama0007@student.monash.edu}{\nolinkurl{dama0007@student.monash.edu}}} \\
   \And
    Dianne Cook
   \\
    Department of Econometrics and Business Statistics \\
    Monash University \\
  Clayton, VIC 3800 \\
  \texttt{\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}} \\
   \And
    Emi Tanaka
   \\
    Department of Econometrics and Business Statistics \\
    Monash University \\
  Clayton, VIC 3800 \\
  \texttt{\href{mailto:emi.tanaka@monash.edu}{\nolinkurl{emi.tanaka@monash.edu}}} \\
   \And
    Kate Hyde
   \\
    Department of Econometrics and Business Statistics \\
    Monash University \\
  Clayton, VIC 3800 \\
  \texttt{} \\
   \And
    Nicholas Tierney
   \\
     \\
   \\
  \texttt{} \\
  }

% Pandoc citation processing

\usepackage{tcolorbox}
\usepackage{fontawesome}
\usepackage{color, colortbl}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\maketitle

\def\tightlist{}

@@

``Open data'' is data that is freely accessible, modifiable, and shareable by anyone for any purpose (Open Knowledge Foundation 2021). This type of data can be useful as example data in statistical textbooks and for research purposes. However, open data is often referred to as what we might call ``wild data'' because it requires substantial cleaning and tidying to tame it into textbook shape. Huebner, Vach, and Cessie (2016) emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building (Huebner et al. 2020).
Data cleaning can be considered to be a part of what is called ``initial data analysis'' (IDA) (Chatfield 1985). In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling. This is also related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data. EDA can be considered to encompass IDA. Dasu and Johnson (2003) say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80\% of the data mining task.

Despite it's importance, this IDA stage is often undervalued and neglected (Chatfield 1985). There are few research papers that document the data cleaning (Wickham 2014). Furthermore, the decisions made in this stage often go unreported in the sense that IDA is often performed in an unplanned and unstructured way and is only shared among restricted parties (Huebner et al. 2020).

This paper demonstrates the steps of IDA, and documents the process, for a prominent open data source, National Longitudinal Survey of Youth 1979, henceforth referred to NLSY79. This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century (Pergamit et al. 2001). In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research (Pergamit et al. 2001; Cooksey 2017). According to Cooksey (2017), thousands of articles, and hundreds of book chapters and monographs have utilized this data. Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data (Pergamit et al. 2001).

Singer and Willett (2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
Our aim is to refresh this textbook data with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences. Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by Singer and Willett (2003). We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the purposes of studying racism, not race (Fullilove 1998).

Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham, Averick, et al. 2019). The data is first organised into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011). (Tidy data shouldn't be confused with ``tame data'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.) The resulting (tame) data is provided in a new R package called \texttt{yowie} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-nlsy79}{}%
Bureau of Labor Statistics, U.S. Department of Labor. 2021. ``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University. Columbus, OH.

\leavevmode\hypertarget{ref-shiny}{}%
Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{shiny: Web Application Framework for R}. \url{https://CRAN.R-project.org/package=shiny}.

\leavevmode\hypertarget{ref-Chatfield1985TIEo}{}%
Chatfield, C. 1985. ``The Initial Examination of Data.'' \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.

\leavevmode\hypertarget{ref-eliznlsy}{}%
Cooksey, Elizabeth C. 2017. ``Using the National Longitudinal Surveys of Youth (Nlsy) to Conduct Life Course Analyses.'' In \emph{Handbook of Life Course Health Development}, edited by Richard M. Lerner Neal Halfon Christoper B. Forrest, 561--77. Cham: Springer. \url{https://doi.org/https://doi.org/10.1007/978-3-319-47143-3_23}.

\leavevmode\hypertarget{ref-DasuTamraparni2003Edma}{}%
Dasu, Tamraparni, and Theodore Johnson. 2003. \emph{Exploratory Data Mining and Data Cleaning}. Wiley Series in Probability and Statistics. Hoboken: WILEY.

\leavevmode\hypertarget{ref-janitor}{}%
Firke, Sam. 2020. \emph{janitor: Simple Tools for Examining and Cleaning Dirty Data}. \url{https://CRAN.R-project.org/package=janitor}.

\leavevmode\hypertarget{ref-racismnotrace}{}%
Fullilove, M. T. 1998. ``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.'' \emph{American Journal of Public Health} 88 (9): 1297--8.

\leavevmode\hypertarget{ref-grimshaw}{}%
Grimshaw, Scott D. 2015. ``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.'' \emph{The American Statistician} 69 (4): 307--14. \url{https://doi.org/10.1080/00031305.2015.1081106}.

\leavevmode\hypertarget{ref-purrr}{}%
Henry, Lionel, and Hadley Wickham. 2020. \emph{purrr: Functional Programming Tools}. \url{https://CRAN.R-project.org/package=purrr}.

\leavevmode\hypertarget{ref-HuebnerMariannePhD2016Asat}{}%
Huebner, Marianne, Werner Vach, and Saskia le Cessie. 2016. ``A Systematic Approach to Initial Data Analysis Is Good Research Practice.'' \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.

\leavevmode\hypertarget{ref-HuebnerMarianne2020Haar}{}%
Huebner, Marianne, Werner Vach, Saskia le Cessie, Carsten Oliver Schmidt, and Lara Lusa. 2020. ``Hidden Analyses: A Review of Reporting Practice and Recommendations for More Transparent Reporting of Initial Data Analyses.'' \emph{BMC Medical Research Methodology} 20 (1): 61--61.

\leavevmode\hypertarget{ref-tamedata}{}%
Kim, A. Y, C. Ismay, and J. Chunn. 2018. ``The Fivethirtyeight R Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.'' \emph{Technology Innovations in Statistics Education} 11 (1). \url{https://doi.org/10.5070/T511103589}.

\leavevmode\hypertarget{ref-KollerManuel2016rARP}{}%
Koller, Manuel. 2016. ``robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models.'' \emph{Journal of Statistical Software} 75 (6): 1--24.

\leavevmode\hypertarget{ref-opendata}{}%
Open Knowledge Foundation. 2021. ``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.'' 2021. \url{http://opendefinition.org/od/2.1/en/}.

\leavevmode\hypertarget{ref-patchwork}{}%
Pedersen, Thomas Lin. 2020. \emph{patchwork: The Composer of Plots}. \url{https://CRAN.R-project.org/package=patchwork}.

\leavevmode\hypertarget{ref-MichaelRPergamit2001DWTN}{}%
Pergamit, Michael R., Charles R. Pierret, Donna S. Rothstein, and Jonathan R. Veum. 2001. ``Data Watch: The National Longitudinal Surveys.'' \emph{The Journal of Economic Perspectives} 15 (2): 239--53.

\leavevmode\hypertarget{ref-R}{}%
R Core Team. 2020. \emph{R: A Language and Environment for Statistical Computing}. Vienna, Austria: R Foundation for Statistical Computing. \url{https://www.R-project.org/}.

\leavevmode\hypertarget{ref-SingerJudithD2003Alda}{}%
Singer, Judith D, and John B Willett. 2003. \emph{Applied Longitudinal Data Analysis: Modeling Change and Event Occurrence}. Oxford u.a: Oxford Univ. Pr.

\leavevmode\hypertarget{ref-brolgar}{}%
Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}. \url{https://github.com/njtierney/brolgar}.

\leavevmode\hypertarget{ref-tukey}{}%
Tukey, John W. (John Wilder). 1977. \emph{Exploratory Data Analysis}. Addison-Wesley Series in Behavioral Science. Reading, Mass.: Addison-Wesley Pub. Co.

\leavevmode\hypertarget{ref-rlm}{}%
UCLA: Statistical Consulting Group. 2021. ``Robust Regression \textbar{} R Data Analysis Examples.'' February 2021. \url{https://stats.idre.ucla.edu/r/dae/robust-regression/}.

\leavevmode\hypertarget{ref-LooMarkvander2018Sdcw}{}%
van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in R}.

\leavevmode\hypertarget{ref-validate}{}%
van der Loo, Mark P. J., and Edwin de Jonge. 2021. ``Data Validation Infrastructure for R.'' \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.

\leavevmode\hypertarget{ref-mass}{}%
Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with S}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.

\leavevmode\hypertarget{ref-tsibble}{}%
Wang, Earo, Dianne Cook, and Rob J Hyndman. 2020. ``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.'' \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78. \url{https://doi.org/10.1080/10618600.2019.1695624}.

\leavevmode\hypertarget{ref-plyr}{}%
Wickham, Hadley. 2011. ``The Split-Apply-Combine Strategy for Data Analysis.'' \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. \url{https://doi.org/10.18637/jss.v040.i01}.

\leavevmode\hypertarget{ref-WickhamHadley2014TD}{}%
---------. 2014. ``Tidy Data.'' \emph{Journal of Statistical Software} 59 (10): 1--23.

\leavevmode\hypertarget{ref-ggplot2}{}%
---------. 2016. \emph{ggplot2: Elegant Graphics for Data Analysis}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.

\leavevmode\hypertarget{ref-stringr}{}%
---------. 2019. \emph{stringr: Simple, Consistent Wrappers for Common String Operations}. \url{https://CRAN.R-project.org/package=stringr}.

\leavevmode\hypertarget{ref-tidyr}{}%
---------. 2020. \emph{tidyr: Tidy Messy Data}. \url{https://CRAN.R-project.org/package=tidyr}.

\leavevmode\hypertarget{ref-tidyverse}{}%
Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019. ``Welcome to the tidyverse.'' \emph{Journal of Open Source Software} 4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.

\leavevmode\hypertarget{ref-dplyr}{}%
Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{dplyr: A Grammar of Data Manipulation}. \url{https://CRAN.R-project.org/package=dplyr}.

\leavevmode\hypertarget{ref-readr}{}%
Wickham, Hadley, and Jim Hester. 2020. \emph{readr: Read Rectangular Text Data}. \url{https://CRAN.R-project.org/package=readr}.

\leavevmode\hypertarget{ref-knitr}{}%
Xie, Yihui. 2014. ``Knitr: A Comprehensive Tool for Reproducible Research in R.'' In \emph{Implementing Reproducible Computational Research}, edited by Victoria Stodden, Friedrich Leisch, and Roger D. Peng. Chapman; Hall/CRC. \url{http://www.crcpress.com/product/isbn/9781466561595}.

\leavevmode\hypertarget{ref-rmarkdown}{}%
Xie, Yihui, Christophe Dervieux, and Emily Riederer. 2020. \emph{R Markdown Cookbook}. Boca Raton, Florida: Chapman; Hall/CRC. \url{https://bookdown.org/yihui/rmarkdown-cookbook}.

\leavevmode\hypertarget{ref-kableExtra}{}%
Zhu, Hao. 2019. \emph{kableExtra: Construct Complex Table with 'kable' and Pipe Syntax}. \url{https://CRAN.R-project.org/package=kableExtra}.

>>>>

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}

@@

\author{
    Dewi Amaliah
   \\
    Department of Econometrics and Business Statistics \\
    Monash University \\
  Clayton, VIC 3800 \\
  \texttt{\href{mailto:dama0007@student.monash.edu}{\nolinkurl{dama0007@student.monash.edu}}} \\
   \And
    Dianne Cook
   \\
    Department of Econometrics and Business Statistics \\
    Monash University \\
  Clayton, VIC 3800 \\
  \texttt{\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}} \\
   \And
    Emi Tanaka
   \\
    Department of Econometrics and Business Statistics \\
    Monash University \\
  Clayton, VIC 3800 \\
  \texttt{\href{mailto:emi.tanaka@monash.edu}{\nolinkurl{emi.tanaka@monash.edu}}} \\
   \And
    Kate Hyde
   \\
    Department of Econometrics and Business Statistics \\
    Monash University \\
  Clayton, VIC 3800 \\
  \texttt{} \\
   \And
    Nicholas Tierney
   \\
    Telethon Kids Institute \\
  Nedlands, WA 6009 \\
  \texttt{} \\
  }

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{tcolorbox}
\usepackage{fontawesome}
\usepackage{color, colortbl}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\begin{document}
\maketitle

@@

Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques. An example of this the wages data made public by Singer and Willett (2003) which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel. The freely available data contains records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S. Department of Labor 2021).

This textbook data was used as an example in Ozlem Ilk's PhD thesis supervised by author Cook (Ilk 2004), and was developed into a case study for use in teaching exploratory data analysis at Iowa State University. The story from modeling the data (and as reported by Singer and Willett) is that wages increase with length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average. The story from an exploratory analysis is very different: the individual experience varies a lot from the overall average. Some individuals experience a decline in wages the longer they are in the workforce, and most experience volatility in their wages.

This disparity between the average and the individual is a part of statistics that is given less attention. It is this disparity which makes this particular data set very important as a textbook example. Textbook data sets have staying power if it has a challenge that has never been completely solved. The iris data (XXX REF) is a prime example, it has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis. (A side note: the iris data is best replaced today with the penguins data (XXX REF), which has similar qualities, is new and does not suffer from a connection with eugenics (Stodel 2020).) The wages data is in this class of textbook data, because it presents this challenge for exploratory data analysis: how can we better summarize and explain the individual experience?

For the field of statistics, and data science by association, it is increasingly important to reach the individual. One might describe this as a divergence of purpose, statistics for public policy or statistics for the public. The two are not the same. As the world becomes more electronically connected combating misinformation and mitigating conspiracy theories require that statistics address the individual. For example, with the wages data, the message for the individual is that you are more than your demographic. The majority of people tracked are not like the average, the average is an anomaly. If you have a bad experience, that your wages have declined over time, you are not alone, there are others like you, and more than you think.

As a textbook data though, the dates are tired. The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published. Teachers of statistics need use contemporary data sets to show how techniques are relevant to today's students. Using tired old textbook data sets sets up a misconception that the field is not current. The wages data is extracted from NSLY79, one of the best examples of open data, which is constantly being updated. It should be possible to continuously refresh the textbook data from the data repository. This is our story of the journey from open (Open Knowledge Foundation 2021) but wild data to textbook data, it's not entirely pretty.

Singer and Willett (2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race. This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century (Pergamit et al. 2001). In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research (Pergamit et al. 2001; Cooksey 2017). According to Cooksey (2017), thousands of articles, and hundreds of book chapters and monographs have utilized this data. Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data (Pergamit et al. 2001).
Our aim is to refresh this textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences. Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by Singer and Willett (2003). We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the \emph{purposes of studying racism, not race} (Fullilove 1998).

This paper demonstrates the steps of cleaning data, and documents the process, as recommended by Huebner, Vach, and Cessie (2016). They emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building. Data cleaning can be considered to be a part of what is called ``initial data analysis'' (IDA) (Chatfield 1985). In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling. This is also related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data. EDA can be considered to encompass IDA. Dasu and Johnson (2003) say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80\% of the data mining task.

Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham et al. 2019). The data is first organised into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011). (Tidy data shouldn't be confused with ``tame data'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.) The resulting (tame) data is provided in a new R package called \texttt{yowie} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-nlsy79}{}}%
Bureau of Labor Statistics, U.S. Department of Labor. 2021. {``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).''} Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University. Columbus, OH.

\leavevmode\vadjust pre{\hypertarget{ref-shiny}{}}%
Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{{shiny: Web Application Framework for R}}. \url{https://CRAN.R-project.org/package=shiny}.

\leavevmode\vadjust pre{\hypertarget{ref-Chatfield1985TIEo}{}}%
Chatfield, C. 1985. {``The Initial Examination of Data.''} \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.

\leavevmode\vadjust pre{\hypertarget{ref-eliznlsy}{}}%
Cooksey, Elizabeth C. 2017. {``Using the National Longitudinal Surveys of Youth (NLSY) to Conduct Life Course Analyses.''} In \emph{Handbook of Life Course Health Development}, edited by Richard M. Lerner Neal Halfon Christoper B. Forrest, 561--77. Cham: Springer. https://doi.org/\url{https://doi.org/10.1007/978-3-319-47143-3_23}.

\leavevmode\vadjust pre{\hypertarget{ref-DasuTamraparni2003Edma}{}}%
Dasu, Tamraparni, and Theodore Johnson. 2003. \emph{Exploratory Data Mining and Data Cleaning}. Wiley Series in Probability and Statistics. Hoboken: WILEY.

\leavevmode\vadjust pre{\hypertarget{ref-janitor}{}}%
Firke, Sam. 2020. \emph{{janitor: Simple Tools for Examining and Cleaning Dirty Data}}. \url{https://CRAN.R-project.org/package=janitor}.

\leavevmode\vadjust pre{\hypertarget{ref-racismnotrace}{}}%
Fullilove, M. T. 1998. {``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.''} \emph{American Journal of Public Health} 88 (9): 1297--98.

\leavevmode\vadjust pre{\hypertarget{ref-grimshaw}{}}%
Grimshaw, Scott D. 2015. {``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.''} \emph{The American Statistician} 69 (4): 307--14. \url{https://doi.org/10.1080/00031305.2015.1081106}.

\leavevmode\vadjust pre{\hypertarget{ref-purrr}{}}%
Henry, Lionel, and Hadley Wickham. 2020. \emph{{purrr: Functional Programming Tools}}. \url{https://CRAN.R-project.org/package=purrr}.

\leavevmode\vadjust pre{\hypertarget{ref-HuebnerMariannePhD2016Asat}{}}%
Huebner, Marianne, Werner Vach, and Saskia le Cessie. 2016. {``A Systematic Approach to Initial Data Analysis Is Good Research Practice.''} \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.

\leavevmode\vadjust pre{\hypertarget{ref-ilk2004}{}}%
Ilk, Ozlem. 2004. {``Exploratory Multivariate Longitudinal Data Analysis and Models for Multivariate Longitudinal Binary Data.''} PhD thesis, Iowa State University. \url{https://doi.org/10.31274/rtd-180813-11012}.

\leavevmode\vadjust pre{\hypertarget{ref-tamedata}{}}%
Kim, A. Y, C. Ismay, and J. Chunn. 2018. {``The Fivethirtyeight {R} Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.''} \emph{Technology Innovations in Statistics Education} 11 (1). \url{https://doi.org/10.5070/T511103589}.

\leavevmode\vadjust pre{\hypertarget{ref-KollerManuel2016rARP}{}}%
Koller, Manuel. 2016. {``{robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models}.''} \emph{Journal of Statistical Software} 75 (6): 1--24.

\leavevmode\vadjust pre{\hypertarget{ref-opendata}{}}%
Open Knowledge Foundation. 2021. {``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.''} 2021. \url{http://opendefinition.org/od/2.1/en/}.

\leavevmode\vadjust pre{\hypertarget{ref-patchwork}{}}%
Pedersen, Thomas Lin. 2020. \emph{{patchwork: The Composer of Plots}}. \url{https://CRAN.R-project.org/package=patchwork}.

\leavevmode\vadjust pre{\hypertarget{ref-MichaelRPergamit2001DWTN}{}}%
Pergamit, Michael R., Charles R. Pierret, Donna S. Rothstein, and Jonathan R. Veum. 2001. {``Data Watch: The National Longitudinal Surveys.''} \emph{The Journal of Economic Perspectives} 15 (2): 239--53.

\leavevmode\vadjust pre{\hypertarget{ref-R}{}}%
R Core Team. 2020. \emph{R: A Language and Environment for Statistical Computing}. Vienna, Austria: R Foundation for Statistical Computing. \url{https://www.R-project.org/}.

\leavevmode\vadjust pre{\hypertarget{ref-SingerJudithD2003Alda}{}}%
Singer, Judith D, and John B Willett. 2003. \emph{Applied Longitudinal Data Analysis: Modeling Change and Event Occurrence}. Oxford u.a: Oxford Univ. Pr.

\leavevmode\vadjust pre{\hypertarget{ref-notiris}{}}%
Stodel, Megan. 2020. {``Stop Using Iris.''} \url{https://www.meganstodel.com/posts/no-to-iris/}.

\leavevmode\vadjust pre{\hypertarget{ref-brolgar}{}}%
Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}}. \url{https://github.com/njtierney/brolgar}.

\leavevmode\vadjust pre{\hypertarget{ref-tukey}{}}%
Tukey, John W. (John Wilder). 1977. \emph{Exploratory Data Analysis}. Addison-Wesley Series in Behavioral Science. Reading, Mass.: Addison-Wesley Pub. Co.

\leavevmode\vadjust pre{\hypertarget{ref-rlm}{}}%
UCLA: Statistical Consulting Group. 2021. {``{Robust Regression \textbar{} R Data Analysis Examples}.''} February 2021. \url{https://stats.idre.ucla.edu/r/dae/robust-regression/}.

\leavevmode\vadjust pre{\hypertarget{ref-validate}{}}%
van der Loo, Mark P. J., and Edwin de Jonge. 2021. {``Data Validation Infrastructure for {R}.''} \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.

\leavevmode\vadjust pre{\hypertarget{ref-LooMarkvander2018Sdcw}{}}%
van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in r}.

\leavevmode\vadjust pre{\hypertarget{ref-mass}{}}%
Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with s}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.

\leavevmode\vadjust pre{\hypertarget{ref-tsibble}{}}%
Wang, Earo, Dianne Cook, and Rob J Hyndman. 2020. {``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.''} \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78. \url{https://doi.org/10.1080/10618600.2019.1695624}.

\leavevmode\vadjust pre{\hypertarget{ref-plyr}{}}%
Wickham, Hadley. 2011. {``The Split-Apply-Combine Strategy for Data Analysis.''} \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. \url{https://doi.org/10.18637/jss.v040.i01}.

\leavevmode\vadjust pre{\hypertarget{ref-WickhamHadley2014TD}{}}%
---------. 2014. {``Tidy Data.''} \emph{Journal of Statistical Software} 59 (10): 1--23.

\leavevmode\vadjust pre{\hypertarget{ref-ggplot2}{}}%
---------. 2016. \emph{{ggplot2: Elegant Graphics for Data Analysis}}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.

\leavevmode\vadjust pre{\hypertarget{ref-stringr}{}}%
---------. 2019. \emph{{stringr: Simple, Consistent Wrappers for Common String Operations}}. \url{https://CRAN.R-project.org/package=stringr}.

\leavevmode\vadjust pre{\hypertarget{ref-tidyr}{}}%
---------. 2020. \emph{{tidyr: Tidy Messy Data}}. \url{https://CRAN.R-project.org/package=tidyr}.

\leavevmode\vadjust pre{\hypertarget{ref-tidyverse}{}}%
Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019. {``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software} 4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.

\leavevmode\vadjust pre{\hypertarget{ref-dplyr}{}}%
Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{{dplyr: A Grammar of Data Manipulation}}. \url{https://CRAN.R-project.org/package=dplyr}.

\leavevmode\vadjust pre{\hypertarget{ref-readr}{}}%
Wickham, Hadley, and Jim Hester. 2020. \emph{{readr: Read Rectangular Text Data}}. \url{https://CRAN.R-project.org/package=readr}.

\leavevmode\vadjust pre{\hypertarget{ref-knitr}{}}%
Xie, Yihui. 2014. {``Knitr: A Comprehensive Tool for Reproducible Research in {R}.''} In \emph{Implementing Reproducible Computational Research}, edited by Victoria Stodden, Friedrich Leisch, and Roger D. Peng. Chapman; Hall/CRC. \url{http://www.crcpress.com/product/isbn/9781466561595}.

\leavevmode\vadjust pre{\hypertarget{ref-rmarkdown}{}}%
Xie, Yihui, Christophe Dervieux, and Emily Riederer. 2020. \emph{R Markdown Cookbook}. Boca Raton, Florida: Chapman; Hall/CRC. \url{https://bookdown.org/yihui/rmarkdown-cookbook}.

\leavevmode\vadjust pre{\hypertarget{ref-kableExtra}{}}%
Zhu, Hao. 2019. \emph{{kableExtra: Construct Complex Table with 'kable' and Pipe Syntax}}. \url{https://CRAN.R-project.org/package=kableExtra}.

\end{CSLReferences}
